Narrative

### 2023-05-20 manuscript/introduction.qmd
intro: updated bias explanations

A positive bias can be expected in heterogeneity estimates of uncorrected effect sizes due to the variance in reliability between primary studies. This is because the attenuation in effect size will then differ from study to study, which results in differences across studies in observed effect sizes beyond their true variability. This is most easily seen by assuming infinite sample size (i.e., no within-study variance) and in the context of a fixed effect model (i.e., with a common underlying mean across studies and no true heterogeneity). Table 1 illustrates how under these conditions three studies with differing reliability will observe variability in effect sizes beyond their true variability (i.e., zero). Table 1 also shows how this effect increases as the true effect size increases.

@@

A negative bias can be expected in heterogeneity estimates of uncorrected effect sizes because given the same unreliability, the attenuation in absolute value increases with effect size. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. As a consequence, in the presence of heterogeneity larger true effect sizes will move further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@

Attenuation of effect size can also be understood as a downwards shift of the distribution of true effect sizes. To illustrate, assume an infinite sample size (i.e., no within-study sampling variance), some heterogeneity $\tau^2$, equal reliability in outcome and measurement such that $R_{xx'} = R_{yy'} = R$, and that all studies within a population of studies have the same reliability. Observed effect sizes given measurement error then corresponds to multiplying all true effect sizes $\theta_i$ by a factor equal to the reliability $R$ (i.e., multiplying the distribution of effect sizes by a constant $R$). If the distribution of effect sizes is normally distributed $\theta_i \sim N(\theta, \tau^2)$ with a mean $\theta$ (i.e., the assumption of the random effects model), then $R \times N(\theta, \tau) = N(R\theta, R^2 \tau^2) = N(R\theta, R\tau)$. Because reliability is always a value \< 1, it can be seen that the square root of the observed heterogeneity for uncorrected effect sizes is less than its true value by a factor $R$. This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately as long as the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large.

To make this concrete, consider an average reliability of 0.8 (the median reported reliability in psychology; Flake et al., 2017). Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. We can consider the positive bias in the same vein. Assuming a common true effect size $\mu$ across studies and infinite sample size within studies we can view $\mu$ as a constant. In this case, attenuation consists of a distribution of reliabilities $R \sim N(\bar{R}, \sigma_R)$ with some mean $\bar{R}$ and standard deviation $\sigma_R$ across studies, that is multiplied by the constant $\mu$ to yield a distribution of observed effect sizes $r \sim N(\mu \bar{R}, \mu \sigma_R)$. To make this concrete, consider some values $\mu = 0.2$, $\bar{R} = 0.8$, $\sigma_R = 0.1$ which then yield $r \sim N(0.16, 0.01)$. Again, this depiction is not entirely correct as reliability is bounded at {0, 1}, but holds approximately as long as the mean reliability $\bar{R}$ and its standard deviation $\sigma_R$ are not too large.

>>>>

A positive bias can be expected in heterogeneity estimates of uncorrected effect sizes due to the variability in reliability between primary studies. This is because the attenuation in effect size will then differ from study to study, which results in differences across studies in observed effect sizes beyond their true variability. This is most easily seen by assuming a common true effect size $\mu$ across studies (i.e., a fixed effect and no true heterogeneity) and infinite sample size within studies. Any observed effect size $\hat{\mu}$ can then be seen as the true effect size $\mu$ multiplied by a study’s reliability $R_i$ (assuming $R_i = R_{xx’} = R_{yy’}$). Hence, the distribution of observed effect sizes is the true effect size $\mu$ multiplied by the distribution of reliabilities across studies. That is, $\hat{mu} \sim \mu \times N(\bar{R}, \sigma_R) = N(\mu \bar{R}, \mu \sigma_R$, where $\bar{R}$ is the average reliability across studies and $\sigma_R$ its standard deviation. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that the observed average effect size will be less than the true effect size by a factor $\bar{R}$ and estimated heterogeneity will be larger than true heterogeneity (i.e., zero) by the product of $\mu \times \sigma_R$ (if heterogeneity is expressed as standard deviation rather than variance). That is, variability in reliability across studies leads to a positive bias in heterogeneity that depends on the size of the true effect size. Table 1 illustrates this effect for three studies with differing reliability and zero true heterogeneity.

@@

A negative bias can be expected in heterogeneity estimates of uncorrected effect sizes because in absolute terms larger effect sizes are more attenuated by a given reliability. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. Consequently, in the presence of heterogeneity attenuation will move larger true effect sizes further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. This negative bias can be observed directly by assuming that an average effect size $\theta$ has some heterogeneity $\tau^2$, all studies have infinite sample size (i.e., no within-study sampling variance), and equal reliability ($R = R_i = R_{xx'} = R_{yy'}$). Any observed effect size $\hat{\theta_i}$ can then be seen as a study’s true effect size $\theta_i$ multiplied by the study reliability $R$. Hence, the distribution of observed effect sizes is the distribution of true effect sizes $N(\theta, \tau^2$ multiplied by the reliability $R$, such that $R \times N(\theta, \tau^2) = N(R \theta, R^2 \tau^2) = N(R \theta, R \tau)$. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that both the observed average effect size and the estimated heterogeneity will be less than their true value by a factor $\bar{R}$ (if heterogeneity is expressed as standard deviation rather than variance). This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately if the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large. Hence, imperfect reliability in studies can be expected to create a negative bias in heterogeneity estimates whenever true heterogeneity is present. To make this concrete, consider an average reliability of 0.8 (the median reported reliability in psychology; Flake et al., 2017). Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@
